Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2024-02-14 manuscript/manuscript_word.Rmd
revision following Dietz's advice and submitted to advances in archaeological practice.

---
title: "Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; cheng.liu@emory.edu]
abstract: "Abstract: This paper presents the Perception-Process-Product (‘Triple P’) framework that aims to expand the scope of experimental archaeology. The Triple P framework emphasizes multi-level variation and interactions across the levels of perception, process, and product to provide a more grounded and richer explanation of the past archaeological record. It consists of three principles: 1) acknowledging the inherent trade-off between control and generalizability in the experimental research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.  \\par \\ Keywords: Experimental archaeology; Ethological analysis; Ethnographic research; Curse of knowledge; Collaborative knowledge production"
date: "`r Sys.Date()`"
output:
  word_document
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
fontsize: 11pt
always_allow_html: true
---

@@

Among the three dimensions of the Triple P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the largest file size compared with the other two. As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files with no size limit and related metadata that can connect with different types of data within the same project. Databrary has three advantages compared with other data storage solutions: a) no cost from the side of researcher; 2) long-term data security monitored by a specialized maintenance team; and 3) fostering potential collaborations between experimental archaeologists and developmental psychologists.

>>>>

---
title: "Variation matters: Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; cheng.liu@emory.edu]
abstract: "Abstract: This paper presents the Perception-Process-Product (\"Triple P\") framework that aims to expand the scope of experimental archaeology. The Triple P framework emphasizes multi-level variation and interactions across the levels of perception, process, and product to provide a more grounded and richer explanation of the past archaeological record. It consists of three principles: 1) acknowledging the inherent trade-off between control and generalizability in the experimental research design; 2) encouraging collaborative projects that involve geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalizes the collection and curation of ethological and ethnographic data in experimental projects.  \\par \\ Keywords: Experimental archaeology; Ethological analysis; Ethnographic research; Curse of knowledge; Collaborative knowledge production"
date: "`r Sys.Date()`"
output:
  word_document
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
fontsize: 11pt
always_allow_html: true
---

@@

Among the three dimensions of the Triple P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the largest file size compared with the other two. As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files with no size limit and related metadata that can connect with different types of data within the same project. Databrary has three advantages compared with other data storage solutions: a) no cost from the side of researchers; 2) long-term data security monitored by a specialized maintenance team; and 3) fostering potential collaborations between experimental archaeologists and developmental psychologists.
